Flaring star-planet interactions in star-planet systems in Kepler and TESS

### 2023-05-11 src/tex/ms.tex
[showyourwork] overleaf sync

% Author list
\author{Ekaterina Ilin}
\author{Katja Poppenh\"ager}

@@

Tidal interaction can also lead to enhanced activity, globally, and locally. Locally, we expect to see periodicity with half the orbital period of the planet, corresponding to the two tidal bulges forming in the stellar envelope~\citep{cuntz2000stellar}. There are neither models to describe how that process leads to enhanced activity, or excess flaring, in star-planet systems, nor are there existing observations. However, in the extreme case of RS CVn binaries, starspots have been observed modulated with half the orbital period of the stellar companion~\citep{ozavci2018recurrent, kriskovics2023ei}. A similar effect was measured when the companion was likely a very low mass star or a brown dwarf~\citep{donati1995activity, frasca2008spots, parks2021interferometric}, and in the case of a K dwarf/white dwarf binary~\citep{hussain2006spot, watson2007roche}. In ellipsoidal binaries, a modulation of flares beyond geometrical effects was observed with Kepler~\citep{gao2016whitelight}, suggesting flaring regions at the location of the extreme tidal bulges in these stars. However, these systems are synchronized, $P_{rot}=P_{orb}$, so that intrinsic active longitudes~\citep{usoskin2007longterm, weber2013theory} cannot be disambiguated from the result of tidal star-star interaction ~\citep{holzwarth2003dynamics}. We can decrease the companion mass further to arrive at local, tidal star-planet interaction, for which, in our sample, spin and orbit are not synchronized. In analogy to~\citet{holzwarth2003dynamics}, we can picture an asymmetry induced in the stellar dynamo by the tidal disturbance, which would lead to a preference for flares at longitudes about $180\,$deg apart from each other.

We cannot tell whether an additional individual flare is produced by tidal or magnetic interaction, but as with magnetic interaction, we can look for deviations from a random distribution of flare times in phase with the relevant period. In the case of tidal interactions, the relevant period is $P_{orb}/2$.

We applied the same technique as described in Section~\ref{sec:methods:adtest} to determine the significance of a local tidal SPI signal. Following~\citet{ilic2022tidal}, we compare the resulting $p$-values with three different models of tidal interaction as proxies for expected excess flaring due to tides in Fig.~\ref{fig:tidal}. We list the derived values in Table~\ref{tab:tidal}.

The first model measures the torque $\partial L / \partial t$, exerted by angular momentum transfer from star to planet or from planet to star, assumin the star has a convective envelope~(\citealt{penev2012constraining}, their Eqns. 1 and 2). We choose the tidal quality factor $Q_*=10^7$. Our results in Fig~\ref{fig:tidal} (left panel) do not depend on the choice of $Q_*$, as we only need to know the tidal torque down to a proportionality factor. We also make the simplifying assumption that the rotation period of the convective envelope is the same as the surface rotation period measured as described in Section~\ref{sec:data:rotationperiods}, ignoring differential rotation effects.

The second model considers the tidal dissipation timescale -- the shorter the timescale, the higher the power of the interaction~\citep{zahn1977tidal}. Following~\citet{ilic2022tidal}, we adopt Eqn.2 from~\citep{albrecht2012obliquities} for the tidal dissipation timescale in the convective envelope. We assume that all stars in our sample have a convective envelope because they flare, and flaring requires a dynamo-generated magnetic field, which can only operate in stars with convective envelopes. Most of our stars are dwarf stars below the Kraft break at spectral type F5, except for KOI-12, which is an F5 sub-giant~\citep{frasca2016activity}.

The third model uses the gravitational perturbation as a proxy for the interaction~\citep{cuntz2000stellar}, originally proposed for star-planet interactions.  Their Eqn. 1 proposes a gravitational perturbation proportional to the mass ratio between planet and star, and $a^{-3}$.

All models require the planetary and stellar mass to be known which we collect from the literature. For missing planetary masses, radii are known, so we can use the mass-radius relations implemented in \texttt{forecaster}~\citep{chen2017probabilistic}. GJ 3323, GJ 674 and GJ 3082 were also missing stellar mass estimates, which we obtained from the \citep{mann2015how, mann2016erratum} relations between absolute Ks magnitude~\citep{skrutskie2006two} and stellar mass for M dwarfs using the distances from \citet{bailer-jones2021estimating}. We list the masses in Table~\ref{tab:tidal}.

>>>>

% Author list
\author{Ekaterina Ilin}
\author{Katja Poppenh\"ager}
\author{Nikoleta Ilic}

@@

Tidal interaction can also lead to enhanced activity, globally, and locally. Locally, we expect to see periodicity with half the orbital period of the planet, corresponding to the two tidal bulges forming in the stellar envelope~\citep{cuntz2000stellar}. There are neither detailed models to describe how that process leads to enhanced activity, or excess flaring, in star-planet systems, nor are there existing observations. However, in the extreme case of RS CVn binaries, starspots often appear modulated with half the orbital period of the system~\citep{olah2002time, ozavci2018recurrent, kriskovics2023ei}. A similar effect was measured when the companion was likely a very low mass star or a brown dwarf~\citep{donati1995activity, frasca2008spots, parks2021interferometric}, and in the case of a white dwarf companion~\citep{hussain2006spot, watson2007roche}. In ellipsoidal binaries, a modulation of flares beyond geometrical effects was observed with Kepler~\citep{gao2016whitelight}, suggesting flaring regions at the location of the extreme tidal bulges in these stars. However, all these systems are synchronized, $P_{rot}=P_{orb}$, so that intrinsic active longitudes~\citep{usoskin2007longterm, weber2013theory, jarvinen2005spots, lanza2009corot} cannot be disambiguated from the result of tidal star-star interaction ~\citep{holzwarth2003dynamics}. In that framework, we can decrease the companion mass to arrive at local, tidal star-planet interaction, for which, in our sample, spin and orbit are not synchronized. In analogy to~\citet{holzwarth2003dynamics}, we can picture an asymmetry induced in the stellar dynamo by the tidal disturbance, which leads to a preference for flares at longitudes about $P_{orb}/2$ apart from each other.

We cannot tell whether an additional individual flare is produced by tidal or magnetic interaction, but we can look for deviations from a random distribution of flare times in phase with the relevant period, i.e., $P_{orb}/2$.

We applied the same technique as described in Section~\ref{sec:methods:adtest} to determine the significance of a local tidal SPI signal. Since we lack a model to relate tidally induced flaring to the systems' properties, we follow~\citet{ilic2022tidal}, and compare the resulting $p$-values with three different models of tidal interaction as proxies for expected excess flaring due to tides in Fig.~\ref{fig:tidal}. We list the derived values in Table~\ref{tab:tidal}.

The first model measures the torque $\partial L / \partial t$, exerted by angular momentum transfer from star to planet or from planet to star, assuming the star has a convective envelope~(\citealt{penev2012constraining}, their Eqns. 1 and 2). We choose the tidal quality factor $Q_*=10^7$. Our results in Fig~\ref{fig:tidal} (left panel) do not depend on the choice of $Q_*$, as we only need to know the tidal torque down to a proportionality factor. We also make the simplifying assumption that the rotation period of the convective envelope is the same as the surface rotation period measured as described in Section~\ref{sec:data:rotationperiods}, ignoring differential rotation effects.

The second model considers the tidal dissipation timescale -- the shorter the timescale, the higher the power of the interaction~\citep{zahn1977tidal}. Following~\citet{ilic2022tidal}, we adopt Eqn. 2 in~\citep{albrecht2012obliquities} for the tidal dissipation timescale in the convective envelope. We assume that all stars in our sample have a convective envelope because they flare, and flaring requires a dynamo-generated magnetic field, which can only operate in stars with convective envelopes. Most of our stars are dwarf stars below the Kraft break at spectral type F5, except for KOI-12, which is an F5 sub-giant~\citep{frasca2016activity}.

The third model uses the gravitational perturbation as a proxy for the interaction~\citep{cuntz2000stellar}, originally derived for star-planet interactions.  Their Eqn. 1 proposes a gravitational perturbation proportional to the mass ratio between planet and star, and $a^{-3}$.

We collect the planetary and stellar masses from the literature~(Table~\ref{tab:tidal}). For missing planetary masses, the radii were known, so used the mass-radius relations implemented in \texttt{forecaster}~\citep{chen2017probabilistic}. GJ 3323, GJ 674 and GJ 3082 were also missing stellar mass estimates, which we obtained from the \citep{mann2015how, mann2016erratum} relations between absolute Ks magnitude~\citep{skrutskie2006two} and stellar mass for M dwarfs using the distances from \citet{bailer-jones2021estimating}. We list the masses in Table~\ref{tab:tidal}.
